Binding-Time Analysis for \textsc{miniKanren}

### 2020-02-28 papers/TEASE-LP/example.tex
TEASE-LP edits

\newcommand{\miniKanren}{\textsc{miniKanren}}

@@

It has been shown in~\cite{lozov2019relational} that online conjunctive partial deduction is capable of improving running time of a program, but we believe that offline specialization may provide better results.
Offline specializers employ some static analysis before the specialization step.
The most prominent of them is \emph{binding-time analysis}.
This analysis is used to determine which data is available statically, and which --- dynamically.
Having this annotations, specialization algorithm can ignore all dynamic data and only symbolically execute a program with respect to the static data.
This usually leads to the more precise and powerful specialization.

@@

@@

A program in miniKanren is a goal accompanied with a set of relation definitions.

The goal of bta is to determine the order in which variables are getting bound within a given goal and all relations called within it.

We will consider a goal in \miniKanren{} to be a disjunction of conjunctions of relations invocations or unifications.
It is always possible to transform any \miniKanren{} goal in this normal form.

We annotate each variable in a goal with either an $undef$ annotation or a natural number.

We start with a goal in which input parameters are annotated with $0$, while all other variables are annotated with $undef$.

Then we run a the following procedure until a fix-point is reached.

Every disjunct is treated seperately from other disjuncts, since they do not influence each other.

Within a conjunction, there can be either a unification or an invocation.
Whenever a new annotation is assigned to some variable use within a conjunction is propagated onto other uses of the same variable.

First we annotate unifications with some partial annotations.
When there are no unifications which can be annotated, we annotate invocations of the current relation.
Then --- all other invocations.
Any of the later annotations can create new partially annotated unifications, so they have to be treated again, in the order they have appeared.

Annotating unification works like this.
There are several possible cases to consider.
The annotations are written in superscripts, $t[x_0, \dots, x_k]$ denotes a term with free variables $x_0, \dots, x_k$.

\begin{itemize}
  \item The unification of an unannotated variable with a term: $x^{undef} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$, where all free variables in a term are annotated with numbers. Then we annotate $x$ with a $1+max\{i_0,\dots,i_k\}$.
  \item The unification of an annotated variable with a term, in which some variables are not annotated: $x^{n} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. Here all unannotated variables within the term are annotated with $1+n$.
  \item The unification of two constructors with the same name and the same number of arguments: $\underline{cons} \ Name \ [t_0^{i_0}, \dots, t_k^{i_k}] \equiv \underline{cons} \ Name \ [s_0^{j_0}, \dots, s_k^{j_k}]$. This is equivalent to the conjunction of the unifications $t_l^{i_l} \equiv s_l^{j_l}$ and is treated accordingly.
  \item Two cases symmetrical to the first two.
  \item If a unification does not conform to any of these cases, do not annotate it, leaving it for later.
\end{itemize}

When annotating invocations, there are possible several cases.

\begin{itemize}
  \item All variables used within an invocation are annotated with $undef$. Annotation in this case is impossible, proceed to the next case.
  \item All variables are annotated. No annotation is needed -- proceed to the next case.
  \item We have already annotated the relation with the same in the consistent direction. Then all $undef$ variables get the annotation $1+n$, where $n$ is the maximum annotation of the annotated variables.
  \item We have not encountered a consistent invocation. Then we try to annotate the body of the corresponding relation.
\end{itemize}

>>>>

\newcommand{\miniKanren}{\textsc{miniKanren}}
\newcommand{\prolog}{\textsc{Prolog}}
\newcommand{\mercury}{\textsc{mercury}}

@@

It has been shown in~\cite{lozov2019relational} that online conjunctive partial deduction is capable of improving running time of a program, but we believe that offline specialization may provide better results.
Offline specializers employ a static analysis before the specialization step.
The most prominent of them is \emph{binding-time analysis}.
It is used to determine which data is available statically, and which --- dynamically.
Having this annotations, specialization algorithm can ignore dynamic data and only symbolically execute a program with respect to the static data.
This usually leads to more precise and powerful specialization.

@@

In the next section we present a binding-time analysis algorithm for \miniKanren{}.

@@

A binding-time analysis starts by annotating a given goal.
Some free variables of the goal are marked with $0$ which denotes they are the \emph{input} or \emph{static} variables, others are annotated with $undef$.
The aim is to determine the order in which the $undef$ variables are bound within the goal as well as to annotate all relation definitions called.
Existing binding-time analysis for pure \prolog{}~\cite{craig2004fully} and \mercury{}~\cite{vanhoof2004binding} only distinguish between static and dynamic variables, which does not help to work out the order.
Natural numbers can serve as annotations~\cite{Thiemann1997AUF} and are better suited to our problem.
Thus we use natural numbers, $0$ and $undef$ for annotations.

We assume that all goals are in \emph{normal form}: a goal is a disjunction of conjunctions of either relation calls or unifications; all free variables are brought up into scope by the $\underline{fresh}$ operator on the top level:
$$G = \underline{fresh} \ V_1 \dots V_n \bigvee \bigwedge (\underline{call} \ R^k \ T_1 \dots T_k \mid T_1 \equiv T_2)$$

The algorithm is to execute the following steps until a fix-point is reached.
Since disjuncts do not influence each other, we treat every disjunct in isolation.
A conjunct is either a unification or a relation call.
At each step, a unification suitable for annotation is selected.
If there are no unifications to annotate, then we select a relation call.
Whenever a new annotation is assigned to a variable, all other uses of the same variable within the conjunction get the same annotation.

Unifications are annotated if there is enough information.
There are several possible cases described below.
We write annotations in superscripts; $t[x_0, \dots, x_k]$ denotes a term with free variables $x_0, \dots, x_k$.
\begin{itemize}
  \item The unification of an unannotated variable with a term in which all free variables are annotated with numbers: $x^{undef} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. In this case we annotate $x$ with $1+max\{i_0,\dots,i_k\}$.
  \item The unification of an annotated variable with a term, in which some variables are not annotated: $x^{n} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. Here all unannotated variables of the term are annotated with $1+n$.
  \item The unification of two constructors with the same name and the same number of arguments: $\underline{cons} \ C^k \ [t_0^{i_0}, \dots, t_k^{i_k}] \equiv \underline{cons} \ C^k \ [s_0^{j_0}, \dots, s_k^{j_k}]$. This is equivalent to the conjunction of the unifications $t_l^{i_l} \equiv s_l^{j_l}$ and is treated accordingly.
  \item Two cases symmetrical to the first two, in which a term is unified with a variable.
\end{itemize}

If a unification does not conform to any of these cases, it should not be annotated at the current step.
Relation calls are only annotated when all unifications have been considered.
A relation call in which all variables have not been annotated (are marked with $undef$) as well as those already annotated with some positive numbers do not need to be considered.
Thus relation calls in which only some of the variables are annotated with positive numbers are to be annotated.

We say that two relation calls have a \emph{consistent} direction if they call a relation with the same name and

\begin{itemize}
  \item
   We have already annotated the relation with the same name in the consistent direction. Then all $undef$ variables get the annotation $1+n$, where $n$ is the maximum annotation of the annotated variables.
  \item We have not encountered a consistent invocation. Then we try to annotate the body of the corresponding relation.
\end{itemize}
